---
title: "Live coding exercise - Pivoting"
format:
  html:
    embed-resources: true
editor: visual
editor_options: 
  chunk_output_type: console
execute:
  echo: true
  warning: false
bibliography: references.bib
---

```{r}
library(tidyverse)
library(gapminder)
```

# tidyr: pivoting

## Example: Waste characterisation data

The data for this exercise was generated as part of a course taught at ETH Zurich, Switzerland [@benaleya2022].

The dataset contains the results of a waste characterisation study of public bins in Zurich, Switzerland. Waste was separated and weighed in kg for the following categories:

-   paper
-   PET
-   metals
-   glass
-   others

Additional metadata on location and type of bin is available.

```{r}
read_csv2(here::here("data/rbtl_fs22_waste_composition_bellevue.csv")) #read_csv lo lee separado por , el read_csv2 lo lee separado por ; 
waste_data <- read_delim(here::here("data/rbtl_fs22_waste_composition_bellevue.csv"), delim = ";") #decimos como queremos que lea el dato separado por ; 




```

```{r}

waste_data_long <- waste_data |> 
  select(number, starts_with("weight")) |>  #seleccionamos todas las que empicen con weight
  rename_with(~str_remove(.x, "weight_")) |>  #renombra las columnas y elimina la parte del weight_ para todas las variables, en este caso .x 
  select(-total) |>  # no nos interesa la de total
  pivot_longer(cols = !number,
               names_to = "weight_category",
               values_to = "weight") #cambiamos y pivoteamos todas menos la de number

```

## Example: Gapminder

The following exercise is taken from [@zimmerman2019].

```{r}

gapminder_wide <- read_csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/main/episodes/data/gapminder_wide.csv")

```

```{r}
gapminder_wide

```

### From wide to long

```{r}
gapminder_long <- gapminder_wide |> 
  pivot_longer(cols = gdpPercap_1952:pop_2007,
               names_to = "obstype_year",
               values_to = "obs_value") |> 
  separate(col = obstype_year, into = c("obs_type", "year"), sep = "_") #los separa en dos columnas


#El siguiente codigo lo que hace es simular que tengamos las variables juntas sin ningun caracter especifico que separe gdpPerCapYYYY so in this usamos la funcion "sep" para separarlo de los ultimos 4 caracteres finales, pero solo funciona para las variables que tengan esa cantidad de caracteres "gdpPerCapYYY". porque si es mayor o menor entonces tendriamos que modificarlo dependiendo del tamaño del texto. 

#gapminder_long |> 
#  unite(col = "obstype_year", obs_type:year, sep = "") |> 
#  separate(col = obstype_year, into = c("obs_type", "year"), sep = -4) |> 
#  count(obs_type)
```

**Your turn**

1.  Using `gapminder_long`, calculate the mean life expectancy, population, and gdpPercap for each continent. **Hint:** use the `group_by()` and `summarize()` functions we learned in the `dplyr` lesson.
2.  Assign the created dataframe to an object with the name `gapminder_long_summary`

```{r}

gapminder_long_summary <- gapminder_long |> 
  group_by(continent, obs_type) |> 
  summarise(Average = mean(obs_value))

gapminder_long_summary |> 
  pivot_wider(names_from = obs_type,
              values_from = Average) |> 
  knitr::kable(digits = 1)
```

### From long to intermediate format

```{r}

gapminder_normal <- gapminder_long |> 
  pivot_wider(names_from = obs_type,
              values_from = obs_value) |> 
  select(country, continent, year, lifeExp, pop, gdpPercap) |> 
  arrange(country, year)
```

# References
